Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2021-12-21 analysis/paper/paper.Rmd
Made png versions of the figures
Changing supplementary material for peer review

* Multivariate exploratory analysis of Mesolithic assemblages in south-eastern Norway
* Explores relevance of artefact categories established in Norwegian archaeology 
* Explores variables associated with mobility patterns in lithic assemblage studies
* Draws on the Whole Assemblage Behavioural Indicators (WABI)
* Relevance for Mesolithic Norway supports the notion that WABI are widely applicable

@@

```{r libraries, eval = TRUE, results = 'hide'}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(knitr)
library(kableExtra)
library(FactoMineR)
library(factoextra)
library(networkD3)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

```{r get-functions, include = FALSE}
source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())

@@

```{r get-data, eval = TRUE, results = 'hide'}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
radiocarbon <- read.csv(here::here('analysis/data/raw_data/radiocarbon_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))

@@

# gis <- read.csv(here::here('../../site_data.csv'))
# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
# gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
```

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100) %>% 
  arrange(mean_date) %>% 
  tibble::rownames_to_column(var = "no")

@@

```{r c14-modelling, cache = TRUE}
# define function to model radiocarbon dates. Only assumption is that 
# dates from a single site belong to a single group
model_rcarb <- function(data){

@@

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9400--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@damlien2018b]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c]. Around the transition to the subsequent Middle Mesolithic or Microlith Phase at c. 8200 BCE, pervasive changes in blade and axe technology occur [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. The next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns [e.g. @jaksland2001; @lindblom1984]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of new flint projectiles [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021, cf. @solheim2018a; @solheim2020a].

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon age determinations are given as the sum of the posterior density esimates. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 13, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

# Assemble plot of date ranges
temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)), label = no)) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = no, colour = excavation_project.x,
                   fill = excavation_project.x), size = 1.8) +
  geom_ridgeline(aes(x = date_grid, height = probabilities * 150,
                     fill = excavation_project.y, 
                     colour = excavation_project.x)) +
  scale_fill_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  scale_colour_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  labs(title = "B", fill = "Excavation project") +
  xlab("Cal. BCE") +
  theme_classic() +
  theme(axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "bottom") +
  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
         colour = FALSE)

@@

The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. The artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading.

@@

As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

@@

The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at <https://osf.io/ehjfc/>, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].

The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table \@ref(tab:tab2)). Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators. Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R. The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021]. The resulting posterior density estimates were then summed for each site. Radiocarbon data are provided in the supplementary data [see also @solheim2020a].

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. This for example pertains to flint tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

@@

```{r lithics}
lithics <- sites %>%
  # Combine variables
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
    small_tool_other = knife_other + scraper_other + drill_other + 
      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
      flake_retouch_other,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

flint <- lithics %>% select(!(ends_with("other")), -grinding_slab_stone, -hammerstone)
```

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120].

@@

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools and debitage from their production characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.

```{r cor1, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10, eval = FALSE}
# Plot with site names tags (probably most immediately relevant for an audience 
# familiar with the sites)
fviz_ca_row(calithics, col.row = "black",
            fill = lithics$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom") 
```

```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot.", fig.height = 13}
# Plot sites
plt1 <- fviz_ca_row(calithics, col.row = "black",
                    fill = lithics$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

blanks_lithics <- c(blanks_flint, "flake_other", "fragment_other", 
                     "blade_other",  "microblade_other")

@@

secondary_lithics <- c(secondary_flint, "axe_other", "axe_debitage_other",
                       "chisel_other", "club_hatchet_other", 
                       "grinding_slab_stone", "projectile_other", 
                       "knife_other", "scraper_other", "drill_other", 
                       "microblade_retouch_other", "blade_retouch_other",
                       "fragment_retouch_other", "flake_retouch_other")

lithics2 <- sites %>% 
  select(-beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is to 
  # retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, flake, flake_other, fragment,
            fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("no")

>>>>

* Multivariate exploratory analysis of Mesolithic assemblages in south-eastern Norway
* Explores relevance of established artefact categories in Norwegian archaeology 
* Explores variables associated with mobility patterns in lithic assemblage studies
* Draws on the Whole Assemblage Behavioural Indicators (WABI)
* Relevance for Mesolithic Norway supports the notion that WABI are widely applicable

@@

```{r libraries, results = 'hide'}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(knitr)
library(kableExtra)
library(FactoMineR)
library(factoextra)
library(networkD3)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

```{r get-functions, include = FALSE}

# Load custom plot functions
source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())

@@

```{r get-data, results = 'hide'}
# Read in data
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
radiocarbon <- read.csv(here::here('analysis/data/raw_data/radiocarbon_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))

@@

@@

```{r wrangling}

# Some initial data preparation
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_nonflint)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_nonflint:flake_retouch_nonflint))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100) %>% 
  arrange(mean_date) %>% 
  tibble::rownames_to_column(var = "no")

@@

```{r c14-modelling, cache = TRUE}
# Define function to model radiocarbon dates. Only assumption is that 
# dates from a single site belong to a single group
model_rcarb <- function(data){

@@

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9300--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@damlien2018b]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; @nærøy2018; but see @viken2018c]. Around the transition to the subsequent Middle Mesolithic or Microlith Phase at c. 8200 BCE, pervasive changes in blade and axe technology occur [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. The next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns [e.g. @jaksland2001; @lindblom1984]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of new flint projectiles [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021; cf. @solheim2020a; @solheim2018a].

@@

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon age determinations are given as the sum of the posterior density esimates. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports. Site numbers match those provided in Table 2.", fig.height = 13, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

# Assemble plot of date ranges
temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)), label = no)) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = no, colour = excavation_project.x,
                   fill = excavation_project.x), size = 1.8) +
  geom_ridgeline(aes(x = date_grid, height = probabilities * 150,
                     fill = excavation_project.y, 
                     colour = excavation_project.x)) +
  scale_fill_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  scale_colour_manual(values = project_colors,
                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
  labs(title = "B", fill = "Excavation project",
       x = "cal BCE", y = "Site number") +
  theme_classic() +
  theme(axis.ticks.y = element_blank(),
        legend.position = "bottom") +
  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
         colour = FALSE)

@@

The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` variables representing different debitage and tool types. The artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g. @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020]. Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g. @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g. @fuglestvedt2007; @eigeland2015; @mansrud2016]. These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application]. In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)) . This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis. Column on the left shows the data as originally compiled, the columns on the right shows how these have been aggregated for the analysis.", fig.height = 7}
# Most of the code for assembling the Sankey plot has been taken directly from 
# various walkthroughs at https://www.r-graph-gallery.com

# Create two columns: old and new variable names
links <- data.frame(value = 1,
        categories = names(sites %>% select(core:flake_retouch_nonflint))) %>% 
  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point",
                                 "transverse_point", "single_edged_point", 
                                 "hognipen_point", "projectile_undefined"), 
                                 "projectile",
                          ifelse(categories %in% c("scraper", "knife", "burin",
                                 "drill", "blade_retouch", "microblade_retouch",
                                 "chip_retouch", "fragment_retouch", 
                                 "flake_retouch"), "small_tool",
                          ifelse(categories %in% c("axe", "chisel"),
                                 "macro_tool", 
                          ifelse(categories %in% c("axe_nonflint", "chisel_nonflint",
                                 "club_hatchet_nonflint"), "macro_tool_nonflint", 
                          ifelse(categories %in% c("knife_nonflint", 
                                 "scraper_nonflint", "drill_nonflint", 
                                 "microblade_retouch_nonflint", 
                                 "blade_retouch_nonflint", 
                                 "fragment_retouch_nonflint",
                                 "flake_retouch_nonflint"), "small_tool_nonflint",
                                 paste0(categories, " ")))))))

# Specify colours for the plot (this can be assembled using a combination of 
# e.g. unique(links$new_connection), rep() and paste(), but I couldn't be 
# bothered so I just copy+pasted and set it up manually).
my_colour <-'d3.scaleOrdinal() .domain(["core", "core_fragment", "blade", "microblade", "chip", "fragment", "flake", "axe_debitage", "macro_tool", "projectile", "microburin", "small_tool", "beach_flint_flint_nodule", "macro_tool_nonflint", "axe_debitage_nonflint", "grinding_slab_stone", "projectile_nonflint", "small_tool_nonflint", "hammerstone", "flake_nonflint", "fragment_nonflint", "chip_nonflint", "blade_nonflint", "microblade_nonflint", "core_nonflint", "core_fragment_nonflint,"]) .range(["lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "#ff8086", "#79d49f", "lightgrey", "#8eb8fa", "lightgrey", "#a88c7d", "lightgrey", "lightgrey", "lightgrey", "#ffee8c",  "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey"])'

@@

As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns [e.g. @Bicho2020; @kitchel2021] and are of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

@@

The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported upon, all data and employed R programming scripts [@rcoreteam] are freely available as a research compendium at <https://osf.io/ehjfc/>, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].

The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table \@ref(tab:tab2)). Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators. Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R. The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021]. The resulting posterior density estimates were then summed for each site.

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. Some additional configurations and ways to aggregate the variables are also available as supplementary material to the paper.

@@

```{r lithics}
lithics <- sites %>%
  # Combine variables
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_nonflint = axe_nonflint + chisel_nonflint + club_hatchet_nonflint,
    small_tool_nonflint = knife_nonflint + scraper_nonflint + drill_nonflint + 
      microblade_retouch_nonflint + blade_retouch_nonflint + fragment_retouch_nonflint +
      flake_retouch_nonflint,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "no") %>% 
  # Deselect a few of the columns not be used
  select(-excavation_project, -excavated_m3, -dating_method)

flint <- lithics %>% select(!(ends_with("nonflint")), -grinding_slab_stone, -hammerstone)
```

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance also means that the structure of the data is

@@

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore performed, excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools and debitage from their production characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.

```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot. Points close together are more similar. By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables. As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns can be made (see Lockyear 2000 for a pedagogical case study).", fig.height = 13}
# Plot sites
plt1 <- fviz_ca_row(calithics, col.row = "black",
                    fill = lithics$mean_date / 1000, title = "A",
                     repel = TRUE, shape = 21, pointsize = 2.9) +
  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

@@

blanks_lithics <- c(blanks_flint, "flake_nonflint", "fragment_nonflint", 
                     "blade_nonflint",  "microblade_nonflint")

@@

secondary_lithics <- c(secondary_flint, "axe_nonflint", "axe_debitage_nonflint",
                       "chisel_nonflint", "club_hatchet_nonflint", 
                       "grinding_slab_stone", "projectile_nonflint", 
                       "knife_nonflint", "scraper_nonflint", "drill_nonflint", 
                       "microblade_retouch_nonflint", "blade_retouch_nonflint",
                       "fragment_retouch_nonflint", "flake_retouch_nonflint")

lithics2 <- sites %>% 
  select(-beach_flint_flint_nodule, -hammerstone, -excavation_project,
          -askeladden_id) %>% 
  # The reason for not using .keep below is to 
  # retain artefact_count
  mutate(chip_prop =  sum(chip, chip_nonflint) / artefact_count,
         core_prop = sum(core, core_nonflint) / artefact_count,
         blanks_prop = sum(blade, microblade, flake, flake_nonflint, fragment,
            fragment_nonflint, blade_nonflint,  microblade_nonflint) / artefact_count,
         secondary_prop =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_nonflint,
                axe_debitage_nonflint, chisel_nonflint, club_hatchet_nonflint, 
                grinding_slab_stone, projectile_nonflint, knife_nonflint,
                scraper_nonflint, drill_nonflint, microblade_retouch_nonflint, 
                blade_retouch_nonflint, fragment_retouch_nonflint,
                flake_retouch_nonflint) / artefact_count,
         nonflint_prop = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_nonflint, -core_nonflint, -core_fragment_nonflint) %>% 
  column_to_rownames("no")
